---
title: "EV Infrastructure Analysis - Exploratory Data Analysis"
output: html_notebook
---

import libraries

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(Hmisc)
library(ggplot2)
```

import data from preprocessing

```{r}
ev_data_by_state <- read_csv("data/ev_data_by_state.csv")
ev_data_by_state

ev_stations_locations <- read_csv("data/ev_station_locations.csv")
ev_stations_locations
```

filter to top 10 states with the most EV stations

```{r}
ev_stations_us_10 <- ev_data_by_state |>
  arrange(desc(total_stations)) |>
  head(10)
ev_stations_us_10
```

visualize top 10 states

```{r}

ggplot(ev_stations_us_10, aes(x = reorder(state, -total_stations), y = total_stations)) + 
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Top 10 States by Total EV Stations",
    x = "State",
    y = "EV Stations"
  ) + 
  geom_text(aes(label = total_stations), vjust = -0.2)
ggsave("outputs/charts/top_10_states_total_ev_stations.png")
```

visualize US charging station locations

```{r}
library(maps)

ggplot(ev_stations_locations, aes(x = longitude, y = latitude)) +
  borders("state") +
  geom_point(alpha = 0.2, color = "green", size = 0.3) +
  coord_fixed(1.3) + 
  labs(title = "US EV Charging Stations")

ggsave("outputs/usa_maps/us_ev_charging_station_map.png")
```

visualize with heat map
show states with largest proportion of EVs to the available charging stations
these states may be underdeveloped and in need of additional EV station rollout
yellow and orange states may need more stations based on the number of vehicles currently registered

```{r}
library(usmap)

plot_usmap(data = ev_data_by_state, values = "evs_per_station", color = "white") + 
  scale_fill_viridis_c(option = "C") +
  labs(title = "EVs Per Charging Station in the US",
       fill = "EVs/Station")

ggsave("outputs/usa_maps/evs_per_charging_station_us.png", bg = "grey")
```

visualize with heat map
show states with the largest proportion of people to the available charging stations
these states have the highest populations relative to the number of stations
yellow and orange states may be in need of more EV stations based on state population

```{r}
plot_usmap(data = ev_data_by_state, values = "people_per_station", color = "white") + 
  scale_fill_viridis_c(option = "C") +
  labs(title = "People Per Charging Station in the US",
       fill = "People/Station")

ggsave("outputs/usa_maps/people_per_charging_station_us.png", bg = "grey")
```

visualize with heat map
show states with the largest highest EV saturation (EV registrations / population)
yellow and orange states could be considered the faster-growing markets compared to purple and blue states
as expected, California has the highest saturation of EVs followed by other Pacific Coast states
other states may see jumps in saturation in the coming decades and would require additional EV stations

```{r}
plot_usmap(data = ev_data_by_state, values = "pop_with_ev", color = "white") + 
  scale_fill_viridis_c(option = "C") +
  labs(title = "EV Saturation by Population",
       fill = "EV Reg/Pop")

ggsave("outputs/usa_maps/ev_saturation_by_population.png", bg = "grey")
```