---
title: "EV Infrastructure Analysis - Preprocessing"
output: html_notebook
---

import libraries

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(Hmisc)
library(ggplot2)
```

import US DoE data
EV Station Data: https://afdc.energy.gov/data_download

```{r}
alt_fuel_stations <- read_csv("data/alt_fuel_stations (Jul 7 2025).csv")
alt_fuel_stations
```

clean columns names using janitor

```{r}
alt_fuel_stations_clean <- clean_names(alt_fuel_stations)
alt_fuel_stations_clean
```

filter to only EV stations ("ELEC")

```{r}
ev_stations <- filter(alt_fuel_stations_clean, fuel_type_code == "ELEC")
ev_stations
```

view full list of column names

```{r}
colnames(ev_stations)
```

select columns for study

```{r}
ev_stations_selected <- ev_stations |>
  select(station_name, city, state, zip, country, latitude, longitude, ev_network)
ev_stations_selected
```

check for missing values

```{r}
sum(is.na(ev_stations_selected))
```

filter to only US market

```{r}
ev_stations_us <- filter(ev_stations_selected, country == "US")
ev_stations_us
```

view unique list of charging netwokrs within dataset

```{r}
ev_networks <- ev_stations_us[ ,"ev_network"]
table(ev_networks)
```

filter to only continental US for geospatial analysis
export station location dataset for geospatial modeling

```{r}
ev_stations_geo <- ev_stations_us |>
  filter(!state %in% c("AK", "PR", "HI"))

write.csv(ev_stations_geo, "data/ev_station_locations.csv")
```


group by state to determine total number of stations per state

```{r}
ev_stations_by_state <- ev_stations_us |>
  group_by(state) |>
  summarise(total_stations = n()) |>
  arrange(desc(total_stations))

ev_stations_by_state
```

import data on number of electric vehicle registrations by state
EV Registration Data: https://afdc.energy.gov/data/10962

```{r}
ev_registrations <- read_csv("data/ev_registration_by_state.csv")
ev_registrations <- clean_names(ev_registrations)
ev_registrations
```

drop total and DC rows and transform state names to abbreviations

```{r}
ev_registrations_50 <- ev_registrations[!(ev_registrations$state) %in% c("District of Columbia", "Total"),]
ev_registrations_50$state <- state.abb[match(ev_registrations_50$state, state.name)]
ev_registrations_50
```

compare total number of ev stations to number of registrations by joining on state
drop PR and DC

```{r}
ev_combined <- ev_stations_by_state |>
  left_join(ev_registrations_50)

ev_combined <- ev_combined[!(ev_combined$state) %in% c("PR", "DC"),]
ev_combined
```

sort by registrations, descending

```{r}
ev_combined <- ev_combined |>
  arrange(desc(ev_combined$registration_count))
ev_combined
```

import population information and transform state names to abbreviations
State Population Data: https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html

```{r}
state_populations <- read_csv("data/state_populations.csv")
state_populations$state <- state.abb[match(state_populations$state, state.name)]
state_populations
```

join population data to existing combined dataset

```{r}
ev_combined_pop <- ev_combined |>
  left_join(state_populations)

ev_combined_pop 
```

calculate electric vehicles per station for each state

```{r}
ev_combined_pop <- ev_combined_pop |>
  mutate(evs_per_station = round(registration_count / total_stations, 2)) |>
  arrange(desc(evs_per_station))
ev_combined_pop
```

calculate number of people per station based on population

```{r}
ev_combined_pop <- ev_combined_pop |>
  mutate(people_per_station = round(population / total_stations, 2)) |>
  arrange(desc(people_per_station))
ev_combined_pop
```

calculate percentage of population with an EV 
note: rough estimate as population includes all ages and some owners may have more than one EV

```{r}
ev_combined_pop <- ev_combined_pop |>
  mutate(pop_with_ev = round(registration_count / population, 4)) |>
  arrange(desc(pop_with_ev))
ev_combined_pop
```

export finalized data frame for further analysis
```{r}
write.csv(ev_combined_pop, "data/ev_data_by_state.csv")
```
