---
title: "EV Infrastructure Analysis - Modeling"
output: html_notebook
---

import libraries

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(Hmisc)
library(ggplot2)
library(randomForest)
```

import data from preprocessing

```{r}
ev_data_by_state <- read_csv("data/ev_data_by_state.csv")
ev_data_by_state

ev_stations_locations <- read_csv("data/ev_station_locations.csv")
ev_stations_locations
```

set random seed

```{r}
set.seed(42)
```


predictive linear modeling based on registrations and population
goal: predict the number of stations that each state should have based on registration count and population

```{r}
model_lm = lm(total_stations ~ registration_count + population, data = ev_data_by_state)
summary(model_lm)
```

predict stations based on model
expected stations = stations the model predicts should exist in each state based solely on registrations and population
station gap = difference between expected stations and current stations

```{r}
ev_data_by_state_model <- ev_data_by_state |>
  rename(ev_stations = total_stations,
         ev_registrations = registration_count) |>
  mutate(expected_stations = round(predict(model_lm), 0),
         station_gap = expected_stations - ev_stations) |>
  arrange(desc(station_gap))
ev_data_by_state_model
```

visualize station gap from linear model

```{r}
plot_usmap(data = ev_data_by_state_model, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State (LM w/ Registration + Population)",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )

ggsave("outputs/modeling/ev_charging_station_gaps_by_state_1.png", bg = "skyblue")
```

```{r}
ev_data_by_state_model_10 <- ev_data_by_state_model |>
  head(10)

ggplot(ev_data_by_state_model_10, aes(x = reorder(state, -station_gap), y = station_gap)) + 
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Top 10 Underbuilt EV States (LM w/ Registration + Population)",
    x = "State",
    y = "Station Gap"
  ) + 
  geom_text(aes(label = station_gap), vjust = -0.2)

ggsave("outputs/modeling/top_10_underbuilt_states_1.png")
```
 
I believe it is a fair assumption to say that the average person does not worry about where to fuel up
when driving an internal combustion engine (ICE) vehicle. Unless driving in extremely rural areas, gas
stations are readily available, have enough pumps, and experience minimal downtime. Let's do a better
job predicting how many EVs each state will need based on the current gasoline infrastructure.

Gas Station Data: https://afdc.energy.gov/files/u/data/data_source/10333/10333_gasoline_stations_year.xlsx

Caveat: data only through 2012 (historical trends show a decline in number of gas stations since 1996)

import and clean up data

```{r}
gas_stations <- read.csv("data/2012_gas_stations_us.csv")

gas_stations <- gas_stations |>
  rename(gas_stations = X2012) |>
  rename(state = State)

gas_stations$state <- state.abb[match(gas_stations$state, state.name)]
gas_stations
```

join with state-level data for analysis

```{r}
ev_gas_stations <- ev_data_by_state_model |>
  left_join(gas_stations) |>
  select(state, ev_stations, gas_stations) |>
  mutate(difference = gas_stations - ev_stations) |>
  arrange(desc(difference))
ev_gas_stations
```

according to this data, there are only three states with more EV stations than gas stations (CO, MA, CA)
all other states have more gas stations than EV stations

goal: predict the number of EV stations each state needs but factor in the number of gas stations this time

join the datasets

```{r}
ev_gas_full <- ev_gas_stations |>
  left_join(ev_data_by_state_model) |>
  select(-...1) |>
  arrange(desc(ev_stations))
ev_gas_full
```

implement the model

```{r}
model_lm_gas = lm(ev_stations ~ ev_registrations + population + gas_stations, data = ev_gas_full)
summary(model_lm_gas)
```

number of gas stations is not significant, so it most likely did not alter the model much

let's look at the predictions from this model

```{r}
ev_gas_predicted <- ev_gas_full |>
  mutate(expected_stations = round(predict(model_lm_gas), 0),
         station_gap = expected_stations - ev_stations) |>
  arrange(desc(station_gap))
ev_gas_predicted
```

visualize station gap from the new linear model with gas stations included

```{r}
plot_usmap(data = ev_gas_predicted, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State (LM w/ Registration + Population + Gas Stations)",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )

ggsave("outputs/modeling/ev_charging_station_gaps_by_state_2.png", bg = "skyblue")
```

factor in geographic area to the model

import state geographic data and convert to state abbreviations
State Area Data: https://www.census.gov/geographies/reference-files/2010/geo/state-area.html

```{r}
state_areas <- read.csv("data/state_areas.csv")
state_areas$state <- state.abb[match(state_areas$state, state.name)]
state_areas
```

join with dataset

```{r}
ev_gas_area <- ev_gas_full |>
  left_join(state_areas) |>
  arrange(desc(sq_mi))
ev_gas_area
```

create new linear model factoring in sq_miles (land area for each state)
note: re-running other models with same dataset for analysis

```{r}
model_lm_area <- lm(ev_stations ~ ev_registrations + population + sq_mi, data = ev_gas_area)
summary(model_lm_area)
```

visualize based on combination

```{r}
ev_gas_area_predicted_lm<- ev_gas_area |>
  mutate(expected_stations = round(predict(model_lm_gas_area), 0),
         station_gap = expected_stations - ev_stations) |>
  arrange(desc(station_gap))
ev_gas_area_predicted_lm
```

```{r}
plot_usmap(data = ev_gas_area_predicted_lm, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State (LM w/ Registration + Population + Gas Stations + Area)",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )

ggsave("outputs/modeling/ev_charging_station_gaps_by_state_5.png", bg = "skyblue")
```

including area seemed to reduce the extremes (mainly, California)

```{r}
anova(model_lm, model_lm_area)
```
