---
title: "EV Infrastructure Analysis - Modeling"
output: html_notebook
---

import libraries

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(Hmisc)
library(ggplot2)
```

import data from preprocessing

```{r}
ev_data_by_state <- read_csv("data/ev_data_by_state.csv")
ev_data_by_state

ev_stations_locations <- read_csv("data/ev_station_locations.csv")
ev_stations_locations
```

differentiate column names

```{r}
ev_data_by_state <- ev_data_by_state |>
  rename(ev_stations = total_stations,
         ev_registrations = registration_count)
ev_data_by_state
```

set random seed

```{r}
set.seed(42)
```

I believe it is a fair assumption to say that the average person does not worry about where to fuel up
when driving an internal combustion engine (ICE) vehicle. Unless driving in extremely rural areas, gas
stations are readily available, have enough pumps, and experience minimal downtime. Let's do a better
job predicting how many EVs each state will need based on the current gasoline infrastructure.

Gas Station Data: https://afdc.energy.gov/files/u/data/data_source/10333/10333_gasoline_stations_year.xlsx

Caveat: data only through 2012 (historical trends show a decline in number of gas stations since 1996)

import and clean up data

```{r}
gas_stations <- read.csv("data/2012_gas_stations_us.csv")

gas_stations <- gas_stations |>
  rename(gas_stations = X2012) |>
  rename(state = State)

gas_stations$state <- state.abb[match(gas_stations$state, state.name)]
gas_stations
```

join with state-level data for analysis

```{r}
ev_data_by_state <- ev_data_by_state |>
  left_join(gas_stations) |>
  mutate(ev_gas_difference = gas_stations - ev_stations) |>
  arrange(desc(ev_gas_difference))
ev_data_by_state
```

factor in geographic area to the model

import state geographic data and convert to state abbreviations
State Area Data: https://www.census.gov/geographies/reference-files/2010/geo/state-area.html

```{r}
state_areas <- read.csv("data/state_areas.csv")
state_areas$state <- state.abb[match(state_areas$state, state.name)]
state_areas
```

join the datasets

```{r}
ev_data_by_state <- ev_data_by_state |>
  left_join(state_areas) |>
  arrange(desc(sq_mi))
ev_data_by_state
```

predict the number of stations that each state should have based on various factors

```{r}
model_poisson = glm(ev_stations ~ ev_registrations + population + gas_stations + sq_mi,
                    data = ev_data_by_state,
                    family = poisson(link = "log"))
summary(model_poisson)
```
```{r}
ev_data_by_state <- ev_data_by_state |>
  mutate(
    expected_stations = round(predict(model_poisson, type = "response"), 0),
    station_gap = expected_stations - ev_stations) |>
  arrange(desc(station_gap))
ev_data_by_state
```

visualize station gap from linear model

```{r}
plot_usmap(data = ev_data_by_state, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )

ggsave("outputs/modeling/underbuilt_states_map.png", bg = "skyblue")
```

```{r}
ev_data_by_state_10 <- ev_data_by_state |>
  head(10)

ggplot(ev_data_by_state_10, aes(x = reorder(state, -station_gap), y = station_gap)) + 
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Top 10 Underbuilt EV States",
    x = "State",
    y = "Station Gap"
  ) + 
  geom_text(aes(label = station_gap), vjust = -0.2)

ggsave("outputs/modeling/top_underbuilt_states.png")
```

save top underbuilt states for geospatial analysis

```{r}
number_of_states <- 5

underbuilt_state_list <- ev_data_by_state |>
  arrange(desc(station_gap)) |>
  head(number_of_states) |>
  pull(state)
underbuilt_state_list

write_rds(underbuilt_state_list, "data/underbuilt_states.rds")
```

