---
title: "Electric Vehicle Infrastructure Analysis"
output: html_notebook
---

Executive Summary

The goal of this project is to determine the states that are most in need of electric vehicle stations based on a number of factors.
These factors include the current number of stations, number of electric vehicle registrations, and state population.
Results from this project should identify the states most in need of adding electric vehicle charging stations.

---

import libraries and dataset

```{r}

library(tidyverse)
library(dplyr)
library(janitor)
library(Hmisc)
library(ggplot2)

alt_fuel_stations <- read_csv("alt_fuel_stations (Jul 7 2025).csv")
alt_fuel_stations
```

set random state seed

```{r}
set.seed(42)
```


clean columns names using janitor

```{r}
alt_fuel_stations_clean <- clean_names(alt_fuel_stations)

alt_fuel_stations_clean
```

filter to only EV stations ("ELEC")

```{r}
ev_stations <- filter(alt_fuel_stations_clean, fuel_type_code == "ELEC")

ev_stations
```

take a look at the values

```{r}
summary(ev_stations)
```

select columns for study (simple set at first)
goal: where are the stations, how many of them are there per state

```{r}
ev_stations_selected <- ev_stations |>
  select(station_name, city, state, zip, country, latitude, longitude, ev_network)

ev_stations_selected
```

look at selected columns

```{r}
summary(ev_stations_selected)

```

check for missing values

```{r}
sum(is.na(ev_stations_selected))
```

now, just US market 

```{r}
ev_stations_us <- filter(ev_stations_selected, country == "US")
ev_stations_us
```

look at different networks (checking for Tesla and Electrify America)

```{r}
ev_networks <- ev_stations_us[ ,"ev_network"]
table(ev_networks)
```
breakdown by state

```{r}
ev_stations_by_state <- ev_stations_us |>
  group_by(state) |>
  summarise(total_stations = n()) |>
  arrange(desc(total_stations))

ev_stations_by_state
```

for plotting (top 10 only)

```{r}
ev_stations_by_state_10 <- ev_stations_us |>
  group_by(state) |>
  summarise(total_stations = n()) |>
  arrange(desc(total_stations)) |>
  head(10)
```


time to plot

```{r}

ggplot(ev_stations_by_state_10, aes(x = reorder(state, -total_stations), y = total_stations)) + 
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Top 10 States by Total EV Stations",
    x = "State",
    y = "EV Stations"
  ) + 
  geom_text(aes(label = total_stations), vjust = -0.2)
```

visualize US charging stations

```{r}
library(maps)

ggplot(ev_stations_us, aes(x = longitude, y = latitude)) +
  borders("state") +
  geom_point(alpha = 0.2, color = "green", size = 0.3) +
  coord_fixed(1.3) + 
  labs(title = "US EV Charging Stations")
```

import data on number of vehicle registrations by state

```{r}
ev_registrations <- read_csv("ev_registration_by_state.csv")
ev_registrations <- clean_names(ev_registrations)
ev_registrations
```

drop total and DC rows

```{r}
ev_registrations_50 <- ev_registrations[!(ev_registrations$state) %in% c("District of Columbia", "Total"),]
ev_registrations_50
```

change state names to abbreviations

```{r}
ev_registrations_50$state <- state.abb[match(ev_registrations_50$state, state.name)]
ev_registrations_50
```

compare total number of ev stations to number of registrations by joining on state
drop PR and DC

```{r}
ev_combined <- ev_stations_by_state |>
  left_join(ev_registrations_50)

ev_combined <- ev_combined[!(ev_combined$state) %in% c("PR", "DC"),]

ev_combined
```

sort by registrations, descending

```{r}
ev_combined <- ev_combined |>
  arrange(desc(ev_combined$registration_count))

ev_combined
```

import population information

```{r}
state_populations <- read_csv("state_populations.csv")
state_populations
```

change state names to abbreviations, again

```{r}
state_populations$state <- state.abb[match(state_populations$state, state.name)]
state_populations
```

join population data to existing combined dataset

```{r}
ev_combined_pop <- ev_combined |>
  left_join(state_populations)

ev_combined_pop 
```

begin simple analysis
goal: calculate electric vehicles per station for each state

```{r}
ev_combined_pop <- ev_combined_pop |>
  mutate(evs_per_station = round(registration_count / total_stations, 2)) |>
  arrange(desc(evs_per_station))

ev_combined_pop
```

goal: calculate number of people per station based on population

```{r}
ev_combined_pop <- ev_combined_pop |>
  mutate(people_per_station = round(population / total_stations, 2)) |>
  arrange(desc(people_per_station))

ev_combined_pop
```

goal: calculate percentage of population with an EV (rough estimate as population includes all ages and some owners may have more than one EV)

```{r}
ev_combined_pop <- ev_combined_pop |>
  mutate(pop_with_ev = round(registration_count / population, 4)) |>
  arrange(desc(pop_with_ev))

ev_combined_pop

ev_stats <- ev_combined_pop # for later use
```

visualize with heat map
show states with largest proportion of EVs to the available charging stations
these states may be underdeveloped and in need of additional EV station rollout
yellow and orange states may need more stations based on the number of vehicles currently registered

```{r}
library(usmap)

plot_usmap(data = ev_combined_pop, values = "evs_per_station", color = "white") + 
  scale_fill_viridis_c(option = "C") +
  labs(title = "EVs Per Charging Station in the US",
       fill = "EVs/Station")

ggsave("evs_per_charging_station_us.png", bg = "grey")

```

visualize with heat map
show states with the largest proportion of people to the available charging stations
these states have the highest populations relative to the number of stations
yellow and orange states may be in need of more EV stations based on state population

```{r}
plot_usmap(data = ev_combined_pop, values = "people_per_station", color = "white") + 
  scale_fill_viridis_c(option = "C") +
  labs(title = "People Per Charging Station in the US",
       fill = "People/Station")

ggsave("people_per_charging_station_us.png", bg = "grey")

```

visualize with heat map
show states with the largest highest EV saturation (EV registrations / population)
yellow and orange states could be considered the faster-growing markets compared to purple and blue states
as expected, California has the highest saturation of EVs followed by other Pacific Coast states
other states may see jumps in saturation in the coming decades and would require additional EV stations

```{r}
plot_usmap(data = ev_combined_pop, values = "pop_with_ev", color = "white") + 
  scale_fill_viridis_c(option = "C") +
  labs(title = "EV Saturation by Population",
       fill = "EV Reg / Pop")

ggsave("ev_saturation_by_population.png", bg = "grey")

```

predictive linear modeling based on registrations and population
goal: predict the number of stations that each state should have based on registration count and population

```{r}
model_lm = lm(total_stations ~ registration_count + population, data = ev_combined_pop)
summary(model_lm)
```

predict stations based on model
expected stations = stations the model predicts should exist in each state based solely on registrations and population
station gap = difference between expected stations and current stations

```{r}
ev_combined_pop <- ev_combined_pop |>
  mutate(expected_stations = round(predict(model_lm), 0),
         station_gap = expected_stations - total_stations) |>
  arrange(desc(station_gap))

ev_combined_pop
```

visualize station gap from linear model

```{r}
plot_usmap(data = ev_combined_pop, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )

ggsave("ev_charging_station_gaps_by_state.png", bg = "grey")
```

```{r}
ev_combined_pop_10 <- ev_combined_pop |>
  head(10)

ggplot(ev_combined_pop_10, aes(x = reorder(state, -station_gap), y = station_gap)) + 
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Top 10 Underbuilt EV States",
    x = "State",
    y = "Station Gap"
  ) + 
  geom_text(aes(label = station_gap), vjust = -0.2)

ggsave("top_10_underbuilt_states.png")
```
 
now, let's get crazy

I believe it is a fair assumption to say that the average person does not worry about where to fuel up
when driving an internal combustion engine (ICE) vehicle. Unless driving in extremely rural areas, gas
stations are readily available, have enough bays, and experience minimal downtime. Let's do a better
job predicting how many EVs each state will need based on the current gasoline infrastructure.

source: https://afdc.energy.gov/files/u/data/data_source/10333/10333_gasoline_stations_year.xlsx
downside: data only through 2012 (historical trends show a decline in number of gas stations since 1996)

import and clean up data

```{r}
gas_stations <- read.csv("2012_gas_stations_us.csv")

gas_stations <- gas_stations |>
  rename(gas_stations = X2012) |>
  rename(state = State)

gas_stations$state <- state.abb[match(gas_stations$state, state.name)]


gas_stations
```

join with ev_stats data for analysis

```{r}
ev_gas_stations <- ev_stats |>
  left_join(gas_stations) |>
  select(state, total_stations, gas_stations) |>
  mutate(difference = gas_stations - total_stations) |>
  arrange(desc(difference))

ev_gas_stations
```

according to this data, there are only three states with more EV stations than gas stations (CO, MA, CA)
all other states have more gas stations than EV stations

let's do some math with the gas station data and factor that into the model
goal: predict the number of EV stations each state needs but factor in the number of gas stations this time

```{r}
ev_gas_stats <- ev_stats |>
  left_join(gas_stations)

model_lm_gas = lm(total_stations ~ registration_count + population + gas_stations, data = ev_gas_stats)
summary(model_lm_gas)
```

number of gas stations is not significant, so it most likely did not alter the model much

let's look at the predictions from this model

```{r}
ev_gas_predicted <- ev_gas_stats |>
  mutate(expected_stations = round(predict(model_lm_gas), 0),
         station_gap = expected_stations - total_stations) |>
  arrange(desc(station_gap))

ev_gas_predicted
```

visualize station gap from the new linear model with gas stations included

```{r}
plot_usmap(data = ev_gas_predicted, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )
```

similar results, so let's try a random forest regressor

```{r}
library(randomForest)

model_rfr_gas = randomForest(total_stations ~ registration_count + population + gas_stations, data = ev_gas_stats)
summary(model_rfr_gas)
```

```{r}
ev_gas_predicted_rfr <- ev_gas_stats |>
  mutate(expected_stations = round(predict(model_rfr_gas), 0),
         station_gap = expected_stations - total_stations) |>
  arrange(desc(station_gap))

ev_gas_predicted_rfr
```

random forest considers CA to be vastly overbuilt (which is not true in actuality, but may be compared to other states)

```{r}
plot_usmap(data = ev_gas_predicted_rfr, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )
```

let's try gradient boosting before I add any additional features to the analysis
 
```{r}
library(gbm)

model_gbm_gas = gbm(total_stations ~ registration_count + population + gas_stations, data = ev_gas_stats,
                    distribution = "gaussian", n.trees = 100)
summary(model_gbm_gas)
```

```{r}
ev_gas_predicted_gbm <- ev_gas_stats |>
  mutate(expected_stations = round(predict(model_gbm_gas), 0),
         station_gap = expected_stations - total_stations) |>
  arrange(desc(station_gap))

ev_gas_predicted_gbm
```

```{r}
plot_usmap(data = ev_gas_predicted_gbm, values = "station_gap", color = "white") + 
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Station Gap"
  ) +
  labs(
    title = "EV Charging Station Gaps by State",
    subtitle = "Red = Underbuilt, Blue = Overbuilt"
  )
```

CA seems to be considered vastly overbuilt every time
This may be true based on other states, but I would hardly consider any state to be considered overbuilt in the
current, growing EV market.
To remedy this issue, I would like to factor in geographic area to the models.

let's clean up some of the column names before this gets even more confusing

```{r}
ev_gas <- ev_gas_stats |>
  rename(ev_stations = total_stations,
         ev_registrations = registration_count)

ev_gas
``` 

import state geographic data
source: https://www.census.gov/geographies/reference-files/2010/geo/state-area.html

```{r}
state_areas <- read.csv("state_areas.csv")
state_areas
```


---

Conclusion

This project sought to identify the states in most need of electric vehicle infrastructure build out.  








